---
title: "**Assignment 6**"
output: html_document
---

#### **Topic: Predicting diabetes using Classification Trees, SVM, and Logistic Regression**

The posted article by Yu et al utilized NHANES data from 1999-2004 to predict diabetes and pre-diabetes using Support Vector Machines. You will conduct a similar analysis using data within the NHANES package in R. 

**Goal:** To predict diabetes status using behavioral, demographic, and clinical features, and to conduct a reproducible analysis to build and compare three classification models: a classification tree, a support vector classifier (linear kernel), and logistic regression.

**Feature Information:** There are 10 features and one outcome variable within the dataset.

* `Diabetes`:    Yes, No (Outcome) [Categorical]
* `Age`:         Age in years [Continuous]
* `Race1`:       Self-reported race/ethnicity [Categorical]
* `Education`:   Highest level of education completed [Categorical]
* `HHIncome`:    Household income category [Categorical]
* `Weight`:      Weight in kilograms [Continuous]
* `Height`:      Height in centimeters [Continuous]
* `Pulse`:       Resting pulse rate (beats per minute) [Continuous]
* `BMI`:         Body Mass Index [Continuous]
* `PhysActive`:  Physically active (Yes/No) [Categorical]
* `Smoke100`:    Smoked at least 100 cigarettes lifetime (Yes/No) [Categorical]


#### **0. Data Preparation**

Load the NHANES data using the NHANES R package. 

**Set up needed libraries**

```{r Q0_libraries, message=FALSE, warning=FALSE}
library(NHANES)
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(e1071)
library(kernlab)
library(pROC)

set.seed(123)
```

**Load and restrict dataset**

Restrict the NHANES data to the list of **11 variables** ("Age", "Race1", "Education", "HHIncome", "Weight", "Height", "Pulse", "Diabetes", "BMI", "PhysActive", "Smoke100"). 

```{r Q0_import, message=FALSE}
vars.keep = c("Age", "Race1", "Education", "HHIncome", "Weight",
               "Height", "Pulse", "Diabetes", "BMI", "PhysActive", "Smoke100")

nhanes.df = NHANES |>
  select(all_of(vars.keep)) |>
  distinct() |>       # remove oversampled duplicate rows
  drop_na()           

nhanes.df
```

**Clean dataset**

```{r Q0_clean, message=FALSE}
# Set reference levels for binary variables and confirm factor types for all categorical variables
nhanes.df = nhanes.df |>
  mutate(
    Diabetes   = factor(Diabetes,   levels = c("No", "Yes")),
    PhysActive = factor(PhysActive, levels = c("No", "Yes")),
    Smoke100   = factor(Smoke100,   levels = c("No", "Yes")),
    Race1      = as.factor(Race1),
    Education  = as.factor(Education),
    HHIncome   = as.factor(HHIncome)
  )

nhanes.df
```

**Overview of continuous features**

```{r Q0_descriptives, message=FALSE}
nhanes.df |>
  summarise(across(where(is.numeric),
                   list(Mean = ~round(mean(.), 2),
                        SD   = ~round(sd(.), 2),
                        Min  = ~round(min(.), 2),
                        Max  = ~round(max(.), 2)))) |>
  pivot_longer(everything(),
               names_to  = c("Variable", "Stat"),
               names_sep = "_(?=[^_]+$)") |>
  pivot_wider(names_from  = Stat,
              values_from = value) |>
  knitr::kable(caption = "Table 1: Descriptive Statistics of Continuous Features")
```

**Check outcome balance**

```{r Q0_balance, message=FALSE}
balance.tbl = nhanes.df |>
  count(Diabetes) |>
  mutate(Percent = round(100 * n / sum(n), 1))

balance.tbl |>
  knitr::kable(caption = "Table 2: Frequency and Proportion of Diabetes Outcome")
```

**NOTE:** From **Table 2**, we see an imbalance in the outcome variable, with 88.2% of participants classified as non-diabetic and 11.8% as diabetic. To address this, down-sampling will be applied during cross-validation to prevent models from being biased toward the majority class. 

**Partition the data into training and testing using a 70/30 split**

```{r Q0_split, message=FALSE}
set.seed(123)

train.indices = createDataPartition(y = nhanes.df$Diabetes, p = 0.7, list = FALSE)
train.data    = nhanes.df[ train.indices, ]
test.data     = nhanes.df[-train.indices, ]

split.counts = tibble(
  Dataset = c("Training data", "Testing data"),
  N       = c(nrow(train.data), nrow(test.data)),
  Percent = round(100 * N / nrow(nhanes.df), 1)
)

split.counts |>
  knitr::kable(caption = "Table 3: Number of Observations in Training and Test Sets")
```

**NOTE:** After reviewing the data structure, descriptive statistics, and outcome balance, all 10 features are retained for modeling. No features are excluded based on the following reasoning: 

* `Age` and `BMI` are well-established clinical risk factors for diabetes 
* `Weight` and `Height` capture body composition information that may complement BMI
* `PhysActive` and `Smoke100` are known behavioral risk factors
* `Race1`, `Education`, and `HHIncome` reflect socioeconomic and demographic context associated with structural health disparities in diabetes risk. 
* `Pulse` is retained as a physiological measure linked to cardiovascular health, which is closely associated with diabetes. 

Lastly, after removing duplicate rows introduced by the NHANES survey oversampling design and retain only complete cases (no missing data), 4,040 observations remained which is sufficient sample size to support all three modeling approaches.

---

#### **1. The Comparison of Three Models**

Construct **three prediction models** to predict diabetes using the features from NHANES. You will optimize each model using cross-validation to choose hyperparameters in the *training data* and then compare performance across models. You will use the following three algorithms to create your prediction models:

First, a shared 10-fold cross-validation control is created to use across all three models. *Down-sampling* is applied within each fold to correct for the class imbalance in the outcome. 

```{r Q1_control, message=FALSE}
train.control = trainControl(
  method          = "cv",
  number          = 10,
  sampling        = "down",
  classProbs      = TRUE,    #allow probability predictions needed for ROC evaluation
  savePredictions = "final"
)
```

**a) Classification Tree**

A classification tree that tunes the complexity parameter (`cp`) via 10-fold cross-validation.

```{r Q1a_tree, message=FALSE, fig.width=20, fig.height=14}
set.seed(123)

grid.tree = expand.grid(cp = seq(0.001, 0.3, by = 0.01))

model.tree = train(
  Diabetes ~ .,
  data      = train.data,
  method    = "rpart",
  trControl = train.control,
  tuneGrid  = grid.tree
)

# Best cp
model.tree$bestTune

# CV performance across cp values
plot(model.tree, uniform = TRUE,
     main = "Figure 1: Classification Tree: Accuracy vs. Complexity Parameter (cp)")

# Visualize the final tree
model.tree$finalModel |>
  rpart.plot(type = 4, extra = 104,
             main = "Figure 2: Optimal Classification Tree — Diabetes Prediction")

# Variable importance
varImp(model.tree)

# Cross-validated confusion matrix from training
confusionMatrix(model.tree)
```

**Description:** The optimal complexity parameter was `cp` = 0.001, selected as the value that maximized cross-validated accuracy of 67.27% after down-sampling. As shown in the `cp` tuning plot in _Figure 1_, accuracy decreases steadily as `cp` increases beyond 0.05, indicating that a more complex tree is needed to capture the patterns in the data. In _Figure 2_, the final tree is large with many splits, reflecting the complexity of diabetes prediction across demographic, socioeconomic, and clinical features. For the variable importance, `Age` (importance = 100.0) and `BMI` (importance = 89.9) were the two most important predictors, followed by demographics and other variables. 

**b) Support Vector Classifier (i.e. Support Vector Machine with a linear classifier)**

```{r Q1b_svm, message=FALSE}
set.seed(123)

model.svm = train(
  Diabetes ~ .,
  data       = train.data,
  method     = "svmLinear",
  trControl  = train.control,
  preProcess = c("center", "scale"),
  tuneGrid   = expand.grid(C = seq(0.001, 2, length = 30))
)

# Best cost parameter
model.svm$bestTune

# Plot accuracy vs. cost
plot(model.svm,
     main = "Figure 3: SVM (Linear) - Accuracy vs. Cost Parameter (C)")

# Cross-validated confusion matrix from training
confusionMatrix(model.svm)

# NOTE: varImp() is not computed for svmLinear with down-sampling
# as caret's ROC-based importance is incompatible with this configuration
```

**Description:** The cost parameter C was optimally set to 1.38 via cross-validation, yielding a mean CV accuracy of 72.5% after down-sampling. As shown in _Figure 3_, accuracy is lowest at very small C values (~65.5%) but stabilizes at 71–72% for moderate to large values, indicating the model is robust to the choice of C above a certain threshold. The cross-validated confusion matrix shows the model correctly identified 9.2% of true diabetic cases and 63.3% of non-diabetic cases on average across folds.

**c) Logistic Regression**

A base logistic regression model using all features, evaluated via 10-fold cross-validation.

```{r Q1c_logistic, message=FALSE}
set.seed(123)

model.logistic = train(
  Diabetes ~ .,
  data       = train.data,
  method     = "glm",
  family     = "binomial",
  trControl  = train.control,
  preProcess = c("center", "scale")
)

# Coefficient summary
summary(model.logistic$finalModel)

# Cross-validated performance
model.logistic$results

# Variable importance
varImp(model.logistic)

# Cross-validated confusion matrix from training
confusionMatrix(model.logistic)
```

**Description:** The logistic regression model achieved a mean CV accuracy of 71.86% and a Kappa of 0.2586 after down-sampling. `Age` was the only statistically significant predictor (β = 1.262, p < 0.001), indicating that older age is strongly associated with higher diabetes risk. `Race1Other` was also significant (β = 0.241, p = 0.039). In addition, the variable importance rankings confirmed that `Age` is the important predictor (importance = 100.0). Lastly, the cross-validated confusion matrix shows the model correctly identified 8.9% of true diabetic cases and 63.0% of non-diabetic cases on average across folds.

---

#### **2. Model Selection**

Select an "optimal" model and calculate final evaluation metrics in the test set. ONLY ONE MODEL SHOULD BE APPLIED IN THE TEST SET. What do you conclude about your final model's performance?

**Compare Models on Training Performance**

```{r Q2_comparison, message=FALSE}
results = resamples(list(
  ClassTree = model.tree,
  SVM       = model.svm,
  Logistic  = model.logistic
))

summary(results)

bwplot(results,
       main = "Figure 4: 10-Fold CV Performance of all models")
```

**Description:** As shown in _accuracy and Kappa results_ above and _Figure 4_, SVM achieved the highest mean CV accuracy (72.47%) and Kappa (0.273), followed by logistic regression (71.86%, Kappa = 0.259), while the classification tree performed notably worse (67.26%, Kappa = 0.185). Despite SVM's marginally higher performance, **logistic regression is selected as the final model** because it offers comparable predictive performance while providing interpretable coefficients. This is an important consideration in a health research context where understanding the direction and magnitude of predictor effects can inform clinical decision-making. 

**Final Model Evaluation (Test Set)**

The final logistic regression model is applied once to the held-out test set to report unbiased evaluation metrics.

```{r Q3_evaluation, message=FALSE}
set.seed(123)

final.pred = predict(model.logistic, newdata = test.data)

confusionMatrix(final.pred, test.data$Diabetes, positive = "Yes")
```

```{r Q3_roc, message=FALSE}
final.pred.prob = predict(model.logistic, newdata = test.data, type = "prob")

roc.analysis = roc(
  response  = test.data$Diabetes,
  predictor = final.pred.prob[, "Yes"]
)

plot(
  1 - roc.analysis$specificities,
  roc.analysis$sensitivities,
  type = "l", col = "steelblue", lwd = 2,
  ylab = "Sensitivity",
  xlab = "1 - Specificity",
  main = "Figure 5: ROC Curve — Logistic Regression (Test Set)"
)
abline(a = 0, b = 1, lty = 2)
legend("bottomright",
       legend = paste("AUC =", round(auc(roc.analysis), 3)),
       col = "steelblue", lwd = 2)

cat("Test Set AUC:", round(auc(roc.analysis), 3), "\n")
```

**Description:** In the test set, the logistic regression model achieved an accuracy of 71.68% and a Kappa of 0.2653. The model correctly identified 78.32% of true diabetic cases (sensitivity) and correctly classifying 70.79% of non-diabetic individuals (specificity). In addition, the AUC of 0.806 confirms good overall performance of the model, substantially above the 0.5 baseline of random guessing. Overall, the model demonstrates good discriminative ability for identifying diabetic individuals in a population setting, though the low PPV (26.4%) suggests it is better suited as a screening tool than a diagnostic one.

---

#### **3. Ethical Considerations of Including Race in a Disease Prediction Model**

In this analysis, we've used `Race` as one of the predictors in the model. Briefly discuss the ethical considerations of including race in a disease prediction model. 

**Answer:** Including `Race1` as a predictor in a diabetes prediction model raises important ethical concerns. Unlike behavioral predictors such as `PhysActive` and `Smoke100`, which represent modifiable risk factors that can be targeted through public health interventions, race is not something an individual can change. When a model assigns higher diabetes risk based on race, it provides no actionable pathway for prevention or treatment, but simply labels certain groups as higher risk without offering a means of intervention. This risks encoding and perpetuating systemic health disparities rather than identifying modifiable risk factors, leading to racial bias or racialized predictions that obscure the true social determinants driving health inequities and may disadvantage already marginalized groups in clinical or policy applications. 

#### **AI tools**

Claude.ai was used in this assignment to assist with debugging R code syntax errors and editing text for clarity and conciseness.